Anomaly detection with kernel density estimation on manifolds

### 2022-07-05 paper/paper.tex
Move swiss roll to appendix b

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

\hypertarget{vkde}{%
\section{Kernel density estimation on manifold learning embeddings}\label{vkde}}

@@

\hypertarget{dckde}{%
\section{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}

However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\). Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\). Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation}
where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).

@@

\hypertarget{riemannian-metric-as-variable-bandwidth}{%
\subsubsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}

@@

\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
\section{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}

@@

\hypertarget{swiss-roll-mapping}{%
\subsubsection{Swiss roll mapping}\label{swiss-roll-mapping}}

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}. The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}

\hypertarget{twin-peaks-mapping}{%
\subsubsection{Twin peaks mapping}\label{twin-peaks-mapping}}

For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
\begin{equation}
\label{eq:twinpeak}
\left\{ 
\begin{array}{lcl}
X = X_1, \\
Y = X_2, \\
Z = \sin(\pi X_1) \tanh (3 X_2).
\end{array}
\right.
\end{equation}

@@

\clearpage\appendix

@@

\hypertarget{volume-density-function-move-to-section-2.2.1}{%
\subsection{Volume density function {[}MOVE TO SECTION 2.2.1{]}}\label{volume-density-function-move-to-section-2.2.1}}

For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\). Then \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\). Define the volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\), as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
\]
The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.

\hypertarget{example-of-s1-for-manifold-density-estimation}{%
\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation}{Example of S\^{}1 for manifold density estimation}}\label{example-of-s1-for-manifold-density-estimation}}

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embe dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

\hypertarget{vkde}{%
\section{Kernel density estimate on Riemannian manifold embeddings}\label{vkde}}

@@

\hypertarget{dckde}{%
\subsection{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}

However, in application, we usually find the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\). Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\). Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation}
where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).

@@

\hypertarget{riemannian-metric-estimation}{%
\subsection{Riemannian metric estimation}\label{riemannian-metric-estimation}}

@@

\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
\subsection{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}

@@

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}. For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
\begin{equation}
\label{eq:twinpeak}
\left\{ 
\begin{array}{lcl}
X = X_1, \\
Y = X_2, \\
Z = \sin(\pi X_1) \tanh (3 X_2).
\end{array}
\right.
\end{equation}

@@

\appendix

@@

\hypertarget{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}{%
\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}{Example of S\^{}1 for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}}\label{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}}

@@

\clearpage

\hypertarget{swissrollappe}{%
\section{Appendix: Simulation with swiss roll mapping}\label{swissrollappe}}

In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.

One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}. The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}
